---
title: "MA677 Midterm Project"
usedpackages: 
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{rotating}
author: "Mingrui Du"
date: "2024-03-26"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(
  ggplot2,
  dplyr,
  tidyverse,
  gridExtra,
  reshape2
)
```

## Order Statistics

```{=latex}
$F_{X_{(1)}}(x) = P(X_{(1)} \leq x)$ = P(at lease one of $X_1, X_2 ... X_n$ is $\leq$ x) \\
\begin{align*}
F_{X_{(1)}}(x) &= P(X_{(1)} \leq x) \\
&= P(\text{at least one of } X_1, X_2, \ldots, X_n \text{ is } \leq x) \\
&= 1 - P(X_{(1)} > x) \\
&= 1 - P(X_1 > x, X_2 > x, \ldots, X_n > x) \\
&= P(X_1 > x)P(X_2 > x) \cdots P(X_n > x) \quad &&\text{(by independence)} \\
&= 1 - [P(X_1 > x)]^n \quad &&(X_i \text{ are identically distributed)} \\
&= 1 - [1 - F(x)]^n \\
f_{X_{(1)}}(x) &= \frac{d}{dx} F_{X_{(1)}}(x) \\
&= \frac{d}{dx} \left\{ 1 - [1 - F(x)]^n \right\} \\
&= n[1 - F(x)]^{n-1} f(x) \\
F_{X_{(n)}}(x) &= P(X_{(n)} \leq x) \\
&= P(X_1 \leq x, X_2 \leq x, \ldots, X_n \leq x) \\
&= P(X_1 \leq x)P(X_2 \leq x)\cdots P(X_n \leq x) \\
&= [F(x)]^n \\
f_{X_{(n)}}(x) &= \frac{d}{dx} F_{X_{(n)}}(x) \\
&= \frac{d}{dx} \left\{[F(x)]^n \right\} \\
&= n[F(x)]^{n-1} f(x) \\
\end{align*}
Similarly, for order statistics in between: \\
\begin{align*}
F_{X_{(k)}}(x) &= P(X_{(k)} \leq x ) \\
&= P(X_{(1)}, X_{(2)}, \ldots X_{(k-1)} \leq x, X_{(k)} = x, X_{(k+1)}, \ldots X_{(n)} > x) \\
&= P(X_1 \leq x)P(X_2 \leq x)\cdots P(X_{(k-1)} \leq x) P(X_{k+1} > x) \cdots P(X_n > x) \\
&= \sum_{j=k}^{n} \binom{n}{j} [F(x)]^j [1 - F(x)]^{n-j} \\
f_{X_{(k)}}(X) &= \frac{d}{dx} F_{X_{(k)}}(X) = \frac{n!}{(k-1)!(n-k)!} [F(x)]^{k-1} [1 - F(x)]^{n-k} f(x) 
\end{align*}
```

[Uniform Distribution, X ~ U(0, 1)]

$f_{X_{(1)}}(x) = n(1-\frac{x-a}{b-a})^{n-1}\frac{1}{b-a} = nx^{(1-1)}(1-x)^{n-1} = n(1-x)^{n-1}I_{(0,\infty)}(x)$ 

$X_{(1)}$ ~ Beta(1, n)

$f_{X_{(k)}}(x) = \frac{n!}{(k-1)!(n-k)!} (\frac{x-a}{b-a})^{k-1} (1 - \frac{x-a}{b-a})^{n-k} \frac{1}{b-a} = I_x(k, n-k+1)$ 

$X_{(k)}$ ~ Beta(k, n-k+1)

$f_{X_{(n)}}(x) = n(\frac{x-a}{b-a})^{n-1} \frac{1}{b-a} = nx^{n-1}I_{(0,1)}(x)$

$X_{(n)}$ ~ Beta(n, 1)

```{r}
par(mfrow = c(2, 3))
x = seq(0, 1, length=100)
plot(x, dbeta(x, 1, 5), ylab='density', type ='l', col='purple', main='X_1')
plot(x, dbeta(x, 2, 4), ylab='density', type ='l', col='purple', main='X_2')
plot(x, dbeta(x, 3, 3), ylab='density', type ='l', col='purple', main='X_3')
plot(x, dbeta(x, 4, 2), ylab='density', type ='l', col='purple', main='X_4')
plot(x, dbeta(x, 5, 1), ylab='density', type ='l', col='purple', main='X_n')
```

[Exponential Distribution, X ~ Exp(1)]

$f_{X_{(1)}}(x) = n(e^{-\lambda x})^{n-1}\lambda e^{-\lambda x} = \lambda  n(e^{-\lambda x})^n$

$f_{X_{(k)}}(x) = \frac{n!}{(k-1)!(n-k)!} (1- e^{-\lambda x})^{k-1} (e^{-\lambda x})^{n-k} \lambda e^{-\lambda x}$ 

$f_{X_{(n)}}(x) = n (1- e^{-\lambda x})^{n-1} \lambda e^{-\lambda x}$

```{r}
# plot by simulation
par(mfrow = c(2, 3))
n <- 5
lambda <- 1
pdf <- function(x, k) {
  (factorial(n) / (factorial(k-1) * factorial(n-k))) * (lambda * exp(-lambda * x)) * (1 - exp(-lambda * x))^(k-1) * exp(-lambda * x)^(n-k)
}
x <- seq(0, 5, length = 1000) 
y <- sapply(x, function(x) pdf(x, 1))
plot(x, y, ylab='density', type ='l', col='purple', main='X_1')
y <- sapply(x, function(x) pdf(x, 2))
plot(x, y, ylab='density', type ='l', col='purple', main='X_2')
y <- sapply(x, function(x) pdf(x, 3))
plot(x, y, ylab='density', type ='l', col='purple', main='X_3')
y <- sapply(x, function(x) pdf(x, 4))
plot(x, y, ylab='density', type ='l', col='purple', main='X_4')
y <- sapply(x, function(x) pdf(x, n))
plot(x, y, ylab='density', type ='l', col='purple', main='X_n')
```

[Normal Distribution, X ~ N(0, 1)]

$f_{X_{(1)}}(x) = n(1-\Phi(\frac{x-\mu}{\sigma}))^{n-1}\frac{1}{\sqrt{2\pi}\sigma} e^{-\frac{1}{2}(\frac{x-\mu}{\sigma})^2}$ 

$f_{X_{(k)}}(x) = \frac{n!}{(k-1)!(n-k)!} (\Phi(\frac{x-\mu}{\sigma}))^{k-1} (1 - \Phi(\frac{x-\mu}{\sigma}))^{n-k} \frac{1}{\sqrt{2\pi}\sigma} e^{-\frac{1}{2}(\frac{x-\mu}{\sigma})^2}$ 

$f_{X_{(n)}}(x) = n(\Phi(\frac{x-\mu}{\sigma}))^{n-1} \frac{1}{\sqrt{2\pi}\sigma} e^{-\frac{1}{2}(\frac{x-\mu}{\sigma})^2}$

```{r}
set.seed(123) 
n <- 50 
reps <- 1000 
sim_data <- replicate(reps, {
  sample <- sort(rnorm(n))
  c(X_1 = sample[1], X_2 = sample[2], X_3 = sample[3], X_4 = sample[4], X_n = sample[n])
})

sim_df <- as.data.frame(t(sim_data))
sim_df$rep <- 1:nrow(sim_df)

ggplot(melt(sim_df, id.vars = "rep"), aes(x = value)) +
  geom_density(aes(fill = variable), alpha = 0.5) +
  labs(x = "Value", y = "Density", title = "Density of Order Statistics") +
  scale_fill_brewer(palette = "Set1", name = "Statistic") +
  theme_minimal()

```



## Exponential 

```{=latex}
[CDF]\\
For X $\sim$ Exp($\lambda$)
\begin{align*}
f(x) &= \lambda e^{-\lambda x}, x \geq 0\\
F(x) &= \int_{0}^{\infty} \lambda e^{-\lambda x} \,dx = -\int_{0}^{\infty} e^{-\lambda x} \,d{(-\lambda x)} \\ 
&= -e^{-\lambda x} \bigg|_{0}^{\infty} = 1 -e^{-\lambda x} \text{, x $\geq$ 0} 
\end{align*}
[MGF]\\
\begin{align*}
M_X(t) = E[e^{tX}] = \int_{0}^{\infty} e^{tx} d{(1 - e^{-\lambda x})} = \int_{0}^{\infty} \lambda e^{(t-\lambda)x} dx = \frac{\lambda}{\lambda - t} \quad \text{for} \quad t < \lambda
\end{align*}
[Expectations, Variances]
\begin{align*}
E[X] &= M_X(1) = \frac{d}{dx} \frac{\lambda}{\lambda - t} \bigg|_{t=0} = \frac{\lambda }{(\lambda - t)^2} \bigg|_{t=0}= \frac{1}{\lambda} \\
E[X^2] &= M_X(2) = \frac{d^2}{dx^2} \frac{\lambda}{\lambda - t} \bigg|_{t=0} = \frac{2\lambda }{(\lambda - t)^3} \bigg|_{t=0} = \frac{2}{\lambda^2}\\
Var[X] &= E[X^2] - E[X]^2 = \frac{1}{\lambda ^2}\\
E[X^3] &= M_X(3) = \frac{d^3}{dx^3} \frac{\lambda}{\lambda - t} \bigg|_{t=0} = \frac{2*3\lambda }{(\lambda - t)^3} \bigg|_{t=0} = \frac{6}{\lambda^3}\\
E[X^4] &= M_X(4) = \frac{d^4}{dx^4} \frac{\lambda}{\lambda - t} \bigg|_{t=0} = \frac{6*4\lambda }{(\lambda - t)^4} \bigg|_{t=0} = \frac{24}{\lambda^4}\\
E[X^5] &= M_X(5) = \frac{d^5}{dx^5} \frac{\lambda}{\lambda - t} \bigg|_{t=0} = \frac{24*5\lambda }{(\lambda - t)^5} \bigg|_{t=0} = \frac{120}{\lambda^5}\\
\end{align*}
```

First Moment: Mean

Second Moment: Variance, measures the spread of distribution

Third Moment: Skewness, provides the degree of asymmetry observed in distribution. Here the positive value indicates this curve is heavy in the left, in other words, has a long tail on the right.

Fourth Moment: Kurtosis focuses on peak or flatness of distribution.

Fifth Moment: Provides insight on shape and asymmetry and tail behavior (from ChatGPT).

Exponential distribution is a special case of Gamma Distribution when n = 1.

For $S = X_1 + X_2 + \ldots + X_n$, $X_i\sim Exp(\lambda)$

$M_S(t) = [M_X(t)]^n = (\frac{\lambda}{\lambda - t})^n$

## Irrigation

```{r}
rotation_times <- c(
  21.80086, 23.74087, 24.6675, 22.1376, 21.4186,
  23.80423, 23.11184, 24.23174, 24.826, 21.44181,
  22.09314, 22.96205, 22.27362, 23.23669, 22.05037,
  21.8075, 22.5501, 24.55148, 23.21969, 24.36872,
  24.56083, 23.8828, 21.84536, 21.90287, 21.55993,
  22.91966, 22.74965, 24.86386, 21.56766, 24.81992,
  22.77892, 21.23745, 22.1006, 21.12459, 21.05793
)
data <- data.frame(rotime = rotation_times)
data$speed <- 1320 * 2 * pi/data$rotime
speed_mean <- mean(data$speed)
speed_sd <- sd(data$speed)
speed_se <- speed_sd / sqrt(nrow(data))
t <- qt(0.95, df = nrow(data) - 1)
margin_error <- t * speed_se
speed_lb <- speed_mean - margin_error
speed_ub <- speed_mean + margin_error
cat("mean: ", speed_mean, "\n")
cat("standard error: ", speed_se, "\n")
cat("90% Confidence Interval: (", speed_lb, ",", speed_ub, ")")
```

The average rotation speed has a mean of 364.2. Standard error 3.26 is relatively small compared to mean. Since sample size is small, we calculate t-value to give 90% confidence interval of the speed, (358.6738 , 369.7162), which means 90% of the speed will be in this range. 

## Acknowledgement

Wuge Li explained concept and strategy of *order statistics*; Jinhu Sun helped on *Irrigation*; and thankfulness to my friends as we together discussed, prepared for both exam and this project.
